---
title: "STA 137 HW #2"
author: "Kathy Mo, Hengyuan Liu, Weilin Cheng"
output:
    pdf_document:
      toc: true # table of content true
      fig_caption: true
---

\newpage
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # show code 
  warning = FALSE,  
  message = FALSE
)
```


\begin{itemize}
  \item You may form a group of 3 students and submit one completed homework for the group.
  \item You can put bullet points like this.
\end{itemize}


```{r,include=FALSE}
library("ggplot2")                    
library("GGally")                   
library("corrplot")
library("ggplot2")
library("qqplotr")
library("leaps")
library("readxl")
```

# Homework 2: Due by 11.59pm on Monday, October 10.

# Problem 1
1. Please find attached the data on US crude oil production in the oil USCrude1.xls.
Let Yt the crude oil production at time t: [time 1 is year 1900 and time 119
is year 2018]

```{r}
us <- read_excel('/Users/davidliu/Desktop/Dataset/USCrude1.xls', sheet='Data 1')
#str(us)
head(us)
#us[-(1:2),]
#colnames(us) <- c("date", "oil")
#head(us)
```
(a)Estimate the trend using
(i) a polynomial of degree 6
(ii) loess with span = 0:2; and 
(iii) spline. Plot Y_t and the three trend estimates against time on the same graph. [Create a variable tm=1:119. Please see pages 5-6 in Handout 4 for R commands.].

```{r}
y <- us$oil
tm <- 1:119 ##a new time variable
plot(tm, y, type = "l") ###just to see the trend visually
```


(i) a polynomial of degree 6
```{r}
###fitting the polynomial trend
polytrnd=lm(y~poly(tm,6))
plot(tm, y, type="l", lty=1, xlab="Time", ylab="oil prod", main="Time series with polynominol Trend")
points(tm, polytrnd$fitted.values, type="l", lty=1, col = "red")
```
(ii) loess with span = 0:2; and 
```{r}
####Fitting loess
loesstrnd=loess(y~tm, span = 0.2)
plot(tm, y, type="l", lty=1, xlab="Time", ylab="oil prod", main="Time series with loess")
points(tm, loesstrnd$fitted, type="l", lty=1, col = "red")

```
(iii) spline. Plot Y_t and the three trend estimates against time on the same graph. [Create a variable tm=1:119. Please see pages 5-6 in Handout 4 for R commands.].
```{r}
trend_spline=function(y, lam){ 
  n=length(y)
  p=length(lam)
  rsq=rep(0, p) 
  y=sapply(y,as.numeric) 
  tm=seq(1/n, 1, by=1/n) 
  xx=cbind(tm, tm^2, tm^3) 
  knot=seq(.1, .9, by=.1) 
  m=length(knot)
  for (j in 1:m){ 
    u=pmax(tm-knot[j], 0); 
    u=u^3 
    xx=cbind(xx,u)
  }
  for (i in 1:p){
    if (lam[i]==0){ 
      ytran=log(y)
    }
    else{ 
      ytran=(y^lam[i]-1)/lam[i]
    }
    ft=lm(ytran~xx)
    res=ft$resid; 
    sse=sum(res^2)
    ssto=(n-1)*var(ytran)
    rsq[i]=1-sse/ssto
  }
  ii=which.max(rsq); lamopt=lam[ii] 
  if (lamopt==0){
      ytran=log(y) } else {
    ytran=y^lamopt
  }
  ft=lm(ytran~xx);
  best_ft=step(ft, trace=0)
  fit=best_ft$fitted; res=best_ft$resid
  result=list(ytrans=ytran, fitted=fit, residual=res, rsq=rsq, lamopt=lamopt)
  return(result)
}

splinetrnd=trend_spline(y, 1) ##note: lam = 1 means no transformation
plot(tm, y, type="l", lty=1, xlab="Time", ylab="oil prod", main="Time series with spline trend")
points(tm, splinetrnd$fitted, type="l", lty=1, col = "red")
```
```{r}
plot(tm, y, type="l", lty=1, xlab="Time", ylab="Temp", main="Temp series with poly, loess, spline trend")
points(tm, polytrnd$fitted.values, type="l", lty=1, col = "red")
points(tm, loesstrnd$fitted, type="l", lty=2, col = "blue")
points(tm, splinetrnd$fitted, type="l", lty=4, col = "brown")
legend("bottomright", c("polyfit","loess", "spline"), lty=c(1,2,4), col = c("red", "blue", "brown"))


```





(b) Plot the three estimates of the rough X_t = Y_t mtagainst time (on separate plots), and obtain the three R2values. [R2is deÖned as 1P(Yt^mt)2= P(YtY )2]

```{r}
roughpoly = polytrnd$residuals
roughloess = loesstrnd$residuals
roughspline = splinetrnd$residual
plot.ts(roughpoly)
plot.ts(roughloess)
plot.ts(roughspline)
```

```{r}
#R square poly
sse = sum((roughpoly)^2)
ssto = sum((y-mean(y))^2)
R2 = 1-sse/ssto
R2
```

```{r}
#Poly
sse = sum((roughpoly)^2)
ssto = sum((y-mean(y))^2)
R2 = 1-sse/ssto
R2

#loess
sse = sum((roughloess)^2)
ssto = sum((y-mean(y))^2)
R2 = 1-sse/ssto
R2

#spline
sse = sum((roughspline)^2)
ssto = sum((y-mean(y))^2)
R2 = 1-sse/ssto
R2


```

```{r}
summary(loesstrnd)
summary(polytrnd)
```



c) For each of the three separate estimates the trend, obtain a histogram of the
rough f^Xtg, and the normal probability plot of f^Xtg. Do these graphs indicate
any serious departure from the normality assumption? Explain. [R commands:
"hist" for histogram, "qqnorm" for normal probability plot.]

```{r}
par(mfrow=c(1,2))
hist(roughpoly, freq = F, main = "Histogram of Rough: Poly Trend", xlab = "Rough part")
qqnorm(roughpoly, main = "Normal prob plot of Rough")
qqline(roughpoly)

```

```{r}
par(mfrow=c(1,2))
hist(roughloess, freq = F, main = "Histogram of Rough: Poly Trend", xlab = "Rough part")
qqnorm(roughloess, main = "Normal prob plot of Rough")
qqline(roughloess)

```

```{r}
par(mfrow=c(1,2))
hist(roughspline, freq = F, main = "Histogram of Rough: Poly Trend", xlab = "Rough part")
qqnorm(roughspline, main = "Normal prob plot of Rough")
qqline(roughspline)

```
Explain: 





(d) Comment on the relative merits of the three fits.


# Problem 2

2. Repeat parts (a)-(d) of Problem 1 when the data is transformed to Y't=
log(Yt).
```{r}
y <- log(y)
```
2(a)Estimate the trend using
(i) a polynomial of degree 6
(ii) loess with span = 0:2; and 
(iii) spline. Plot Y_t and the three trend estimates against time on the same graph. [Create a variable tm=1:119. Please see pages 5-6 in Handout 4 for R commands.].

```{r}
tm <- 1:119 ##a new time variable
plot(tm, y, type = "l") ###just to see the trend visually

```



(i) a polynomial of degree 6
```{r}
###fitting the polynomial trend
polytrnd=lm(y~poly(tm,6))
plot(tm, y, type="l", lty=1, xlab="Time", ylab="oil prod", main="Time series with polynominol Trend")
points(tm, polytrnd$fitted.values, type="l", lty=1, col = "red")

```
(ii) loess with span = 0:2; and 
```{r}
####Fitting loess
loesstrnd=loess(y~tm, span = 0.2)
plot(tm, y, type="l", lty=1, xlab="Time", ylab="oil prod", main="Time series with loess")
points(tm, loesstrnd$fitted, type="l", lty=1, col = "red")
```
(iii) spline. Plot Y_t and the three trend estimates against time on the same graph. [Create a variable tm=1:119. Please see pages 5-6 in Handout 4 for R commands.].
```{r}
trend_spline=function(y, lam){ 
  n=length(y)
  p=length(lam)
  rsq=rep(0, p) 
  y=sapply(y,as.numeric) 
  tm=seq(1/n, 1, by=1/n) 
  xx=cbind(tm, tm^2, tm^3) 
  knot=seq(.1, .9, by=.1) 
  m=length(knot)
  for (j in 1:m){ 
    u=pmax(tm-knot[j], 0); 
    u=u^3 
    xx=cbind(xx,u)
  }
  for (i in 1:p){
    if (lam[i]==0){ 
      ytran=log(y)
    }
    else{ 
      ytran=(y^lam[i]-1)/lam[i]
    }
    ft=lm(ytran~xx)
    res=ft$resid; 
    sse=sum(res^2)
    ssto=(n-1)*var(ytran)
    rsq[i]=1-sse/ssto
  }
  ii=which.max(rsq); lamopt=lam[ii] 
  if (lamopt==0){
      ytran=log(y) } else {
    ytran=y^lamopt
  }
  ft=lm(ytran~xx);
  best_ft=step(ft, trace=0)
  fit=best_ft$fitted; res=best_ft$resid
  result=list(ytrans=ytran, fitted=fit, residual=res, rsq=rsq, lamopt=lamopt)
  return(result)
}

splinetrnd=trend_spline(y, 1) ##note: lam = 1 means no transformation
plot(tm, y, type="l", lty=1, xlab="Time", ylab="oil prod", main="Time series with spline trend")
points(tm, splinetrnd$fitted, type="l", lty=1, col = "red")
```
```{r}
plot(tm, y, type="l", lty=1, xlab="Time", ylab="Temp", main="Temp series with poly, loess, spline trend")
points(tm, polytrnd$fitted.values, type="l", lty=1, col = "red")
points(tm, loesstrnd$fitted, type="l", lty=2, col = "blue")
points(tm, splinetrnd$fitted, type="l", lty=4, col = "brown")
legend("bottomright", c("polyfit","loess", "spline"), lty=c(1,2,4), col = c("red", "blue", "brown"))


```

(b) Plot the three estimates of the rough X_t = Y_t mtagainst time (on separate plots), and obtain the three R2values. [R2is deÖned as 1P(Yt^mt)2= P(YtY )2]

```{r}
roughpoly = polytrnd$residuals
roughloess = loesstrnd$residuals
roughspline = splinetrnd$residual
plot.ts(roughpoly)
plot.ts(roughloess)
plot.ts(roughspline)
```

```{r}
#R square poly
sse = sum((roughpoly)^2)
ssto = sum((y-mean(y))^2)
R2 = 1-sse/ssto
R2
```


```{r}
#Poly
sse = sum((roughpoly)^2)
ssto = sum((y-mean(y))^2)
R2 = 1-sse/ssto
R2

#loess
sse = sum((roughloess)^2)
ssto = sum((y-mean(y))^2)
R2 = 1-sse/ssto
R2

#spline
sse = sum((roughspline)^2)
ssto = sum((y-mean(y))^2)
R2 = 1-sse/ssto
R2


```


c) For each of the three separate estimates the trend, obtain a histogram of the
rough f^Xtg, and the normal probability plot of f^Xtg. Do these graphs indicate
any serious departure from the normality assumption? Explain. [R commands:
"hist" for histogram, "qqnorm" for normal probability plot.]

```{r}
par(mfrow=c(1,2))
hist(roughpoly, freq = F, main = "Histogram of Rough: Poly Trend", xlab = "Rough part")
qqnorm(roughpoly, main = "Normal prob plot of Rough")
qqline(roughpoly)

```
```{r}
par(mfrow=c(1,2))
hist(roughloess, freq = F, main = "Histogram of Rough: Poly Trend", xlab = "Rough part")
qqnorm(roughloess, main = "Normal prob plot of Rough")
qqline(roughloess)

```

```{r}
par(mfrow=c(1,2))
hist(roughspline, freq = F, main = "Histogram of Rough: Poly Trend", xlab = "Rough part")
qqnorm(roughspline, main = "Normal prob plot of Rough")
qqline(roughspline)

```
Explain: 





(d) Comment on the relative merits of the three fits.



(e) Based on your analyses, would you prefer to work with the original series,
or the transformed series flog(Yt)g? Explain you answer.






\newpage

### Code Appendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
